The role of Similarity in native spoken word recognition

### 2025-03-22 manuscript/manuscript.qmd
Finish manuscript, appendix, clean repo, and add Docker image

::: {.content-visible when-format="pdf"}

@@

:::

::: {.content-visible when-format="docx"}

One exception is the study by @otake2007interlingual, who analysed 194 instances of Soramimi, broadcasted between 1992 and 2007 by the TV show *Soramimi hour*, in which examples of Soramimi were presented the the audience for comedy purposes. These instances consisted of homophonic translations of English song lyrics to words and phrases in Japanese. The vast majority of the reported instances were phrases (96%) and a few, single words (4%). There was relatively high variability in the degree to which phonetic features of the presented English lyrics were preserved, with some resulting Japanese words or phrases sharing high overlap with the original English strings of sounds, and some sharing very little overlap. An analysis of the few instances of homophonic translations of single words revealed three main phonological processes that could explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /krʌɪ/ to *kurai*, /kuɾai/ [dark]), deletion (e.g., *go* /goʊ/ to *go* /go/ [*Go*, a board game], and alternation (e.g., *low* /loʊ/ to *rou* /ɾoo/ [wax]). This suggests that the reported homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to the Japanese phoneme inventory and Japanese phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic].

In Experiment 1, we collected data from two groups of British English natives living in the United Kingdom. One group was presented with Catalan words, the other with Spanish words. We examined the extent to which participants were able to use the phonological similarity between the presented word (in Catalan or Spanish) and its target translation to provide accurate responses. In Experiment 2, we tested a group of (European) Spanish natives in the same task, who were presented with Catalan words. Catalan and Spanish are two Romance languages whose close typological distance is reflected in the fact that they share many cognates, where English is a Germanic language that shares considerably fewer cognates with Catalan and Spanish. By testing participants translating words from typologically close or distant languages, we expected to widen the range of the phonological similarity scores of the translation pairs involved in the experimental task, therefore allowing us to explore potential cross-language differences in participant’s performance. One unexpected finding was that participants in Experiments 1 and 2 were surprisingly good at translating a subset of words which had low phonological similarity with their correct translation. We were concerned that this may be caused by some prior knowledge of specific words by our participants, as some Spanish words are commonly seen in media or product labels, making them familiar even to monolingual speakers of English. In Experiment 3, we collected additional data from a new group of British English natives. The design was closely modelled after Experiment 1, except that after providing their response in each trial, participants reported whether they had previous knowledge of the presented word. In the final section of this article, we present analyses on the joint data sets of Experiments 1 to 3.
:::

@@

For each word pair in the Catalan and Spanish lists, we defined three predictors of interest: the lexical frequency of the correct English translation (*Frequency*), the phonological similarity between the presented word and its correct English translation (*Similarity*), and the presented word's number of cross-language phonological neighbours (*CLPN*) in English. *Frequency* was included as a nuisance predictor, under the hypothesis that---keeping other predictors constant---participants would provide correct responses more often when the correct translations are high frequency words in their native language. Lexical frequencies of correct translations were extracted from SUBTLEX-UK [@van2014subtlex], and transformed to Zipf scores. Translations without a lexical frequency value were excluded from data analysis (`r stim_exc_freq$cat_eng$n` in the Catalan list, `r stim_exc_freq$spa_eng$n` in the Spanish list).  *Similarity*, our main predictor of interest, was calculated as the Levenshtein similarity between the X-SAMPA transcriptions of each pair of translations using the `stringdist` R package [@van2014stringdist]. The Levenshtein distance computes the edit distance between two character strings---in this case, two phonological transcriptions---by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. We divided this edit distance by the number of characters in the longest X-SAMPA transcription of the translation pair. This transformation accounts for the fact that Levenshtein distance tends to increase with the length of the transcriptions. For interpretability, we subtracted this proportion from one, so that values closer to one correspond to higher similarity between phonological transcriptions, and values closer to zero correspond to lower similarity between phonological transcriptions. For example, the *table* (`teIb@l`)-*mesa* (`mesa`) translation pair had `r scales::percent(stringdist::stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (`trEIn`)-*tren* (`tREn`) translation pair had `r scales::percent(stringdist::stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity. We calculated the number of *CLPN* for each Catalan and Spanish word by counting the number of English words with same or higher lexical frequency, and whose phonological transcription (in X-SAMPA format) differed by up to one phoneme from that of the presented Catalan or Spanish word. Lexical frequencies and phonological transcriptions were extracted from the multilingual database CLEARPOND [@marian2012clearpond]^[Phonological transcriptions in CLEARPOND were generated from eSPEAK, [http://espeak.sourceforge.net/](http://espeak.sourceforge.net/)]]. @tbl-stimuli summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

@@

::: {.content-visible when-format="docx"}
![Eq. 1](_assets/img/model.PNG)
:::

::: {.content-visible when-format="pdf"}
$$
\begin{aligned}
&\textbf{Likelihood}  \\
y_{i} \sim & \text{Bernoulli}(p_{i}) \\ \\
&\textbf{Parameters}  \\
\text{Logit}(p_{i}) = &  \beta_{0[p,w]} + \beta_{1[p]} \text{Frequency}_{i} + \beta_{2[p]} \text{Similarity}_i +  \beta_{3[p]} \text{CLPN}_i + \beta_{4[p]} \text{Group}_i + \\
&\beta_{5[p]} (\text{Similarity}_i \times \text{CLPN}_i) + \beta_{6[p]} (\text{Similarity}_i \times \text{Group}_i) + \\
& \beta_{7[p]} (\text{CLPN}_i \times \text{Group}_i) + \beta_{8[p]} (\text{Similarity}_i \times \text{CLPN}_i \times \text{Group}_i) \\ \\
\beta_{0-8[p,w]} \sim & \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \text{ and  word } w \text{ in 1, ..., } W \\
\beta_{1-8[p]} \sim &  \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \\ \\
&\textbf{Prior}  \\
\mu_{\beta_{p,w}}  \sim &  \mathcal{N}(0, 0.1) \\
\sigma_{\beta_{p}},  \sigma_{\beta_{w}} \sim & \text{HalfCauchy}(0, 0.1) \\
\rho_{p}, \rho_{w} \sim & \text{LKJCorr}(8) \\
\end{aligned}
$$ {#eq-1}

:::

@@

All analyses were performed in the R environment [@rcoreteam2013language]. We used the tidyverse family of R packages [@wickham2019welcome] to process data and to generate figures. We used the `brms` R package [@burkner2017brms] using the `cmdstanr` back-end to the Stan probabilistic language [@carpenter2017stan] to estimate and compare the models (see Appendix 1 for more details on the models). Numeric predictors were standardised before entering the model by subtracting the mean and dividing by the standard deviation.

@@

tbl_data |>
  gt(
    rowname_col = "group",
    groupname_col = "exp"
  ) |>
  fmt_number(is.numeric, decimals = 2) |>
  # gtExtras::gt_plt_dist(correct_list, type = "density") |>
  fmt_integer(c(matches("sum|min|max"), n_participants)) |>
  fmt_number(matches("correct"), scale_by = 100) |>
  tab_spanner("Accuracy (%)", matches("correct")) |>
  tab_spanner("Valid trials", matches("n_trials")) |>
  cols_merge_range(n_trials_min, n_trials_max) |>
  cols_merge_range(correct_min, correct_max) |>
  cols_label(
    n_trials_sum = "N trials",
    n_participants = "N",
    n_trials_mean = "Mean",
    n_trials_sd = "SD",
    n_trials_min = "Range",
    correct_mean = "Mean",
    correct_sd = "SD",
    correct_sem = "SE",
    correct_min = "Range"
  ) |>
  tab_style(
    cell_text(align = "center"),
    cells_column_labels()
  ) |>
  tab_style(
    cell_text(style = "italic"),
    cells_column_labels()
  ) |>
  tab_style(
    cell_text(weight = "bold"),
    cells_column_spanners()
  ) |>
  tab_style(
    cell_text(align = "left"),
    list(cells_body(), cells_column_labels())
  )
```

@@

epreds_1 |> 
  ggplot(aes(lv_std, .epred)) +
  facet_grid(group ~ neigh_n_h_std) +
  stat_lineribbon(aes(fill_ramp = after_stat(level)),
    linewidth = 1 / 2,
    .width = c(0.95, 0.89, 0.78, 0.67, 0.50)
  ) +
  labs(
    x = "Similarity\n(Levenshtein similarity with correct translation)",
    y = "p(Correct)",
    colour = "Cross-language neighbourhood density",
    fill = "Cross-language neighbourhood density",
    linetype = "Cross-language neighbourhood density",
    fill_ramp = "CrI (%)"
  ) +
  scale_x_continuous(
    labels = \(x) scales::percent((x * sd(dataset_1$lv, na.rm = TRUE)) + mean(dataset_1$lv, na.rm = TRUE)), 
    breaks = (seq(0, 1, 0.2) - mean(dataset_1$lv, na.rm = TRUE)) / sd(dataset_1$lv, na.rm = TRUE)) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Reds") +
  theme(
    legend.position = "none",
    legend.box = "vertical"
  )
```

@@

epreds_2 |> 
  ggplot(aes(lv_std, .epred)) +
  facet_grid(~ neigh_n_h_std) +
  stat_lineribbon(aes(fill_ramp = after_stat(level)),
    linewidth = 1 / 2,
    .width = c(0.95, 0.89, 0.78, 0.67, 0.50)
  ) +
  labs(
    x = "Similarity\n(Levenshtein similarity with correct translation)",
    y = "p(Correct)",
    colour = "Cross-language neighbourhood density",
    fill = "Cross-language neighbourhood density",
    linetype = "Cross-language neighbourhood density",
    fill_ramp = "CrI (%)"
  ) +
  scale_x_continuous(
    labels = \(x) scales::percent((x * sd(dataset_2$lv, na.rm = TRUE)) + mean(dataset_2$lv, na.rm = TRUE)), 
    breaks = (seq(0, 1, 0.2) - mean(dataset_2$lv, na.rm = TRUE)) / sd(dataset_2$lv, na.rm = TRUE)) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Reds") +
  theme(
    legend.position = "none",
    legend.box = "vertical"
  )
```

@@

Experiment 2 was an extension of Experiment 1 to a population of monolinguals whose native language is typologically close to the presented language. We presented Catalan words to Spanish native adults who were reportedly unfamiliar with Catalan. Our results indicate a similar pattern of results as those in Experiment 1: participants were able to provide correct translations of presented Catalan words, provided that the Catalan words shared some degree of phonological similarity with their Spanish translation, and that the number of phonological neighbours with higher lexical frequency was reduced. In contrast with the results in Experiment 1, the positive impact of phonological similarity on participants' performance in Experiment 2 larger. Spanish natives in Experiment 2 provided exploited phonological similarity more strongly than English natives in Experiment 1. Overall, this suggests that participants in Experiment 2, who were natives of a typologically similar language (Spanish) to the presented language (Catalan), benefited more strongly from phonological similarity than participants in Experiment 1, who were natives of typologically less similar language (English) to the presented language (Catalan, Spanish). Participants from both Experiment 1 and 2 benefited strongly from phonological similarity to correctly translate words from a non-native, reportedly unfamiliar language. This pattern of results holds for most of the presented stimuli, but some low-similarity Catalan and Spanish words were responded to surprisingly accurately by English listeners. Given that participants were reportedly unfamiliar with both languages, it was expected that participants would be very unlikely to provide correct translations for words sharing little to no phonological similarity to their correct translation. @tbl-surprises shows a list of Catalan and Spanish words to which participants provided responses with $\geq$ 10 average accuracy.

::: {.content-visible when-format="pdf"}

```{r tbl-surprises}
#| label: tbl-surprises
#| tbl-cap: "List of items with unexpectedly high accuracy: the Levenshtein similarity score between the presented word (in Catalan or Spanish) and their correct English translation is zero, but participants, who are reportedly unfamiliar with the presented language, were on average >10% likely to guess the correct translation."
ipa_to_tipa <- function(ipa) {
  dict <- c(
    "ɑ" = "A", 
    "ɒ" = "6",
    "æ" = "\\ae", 
    "ʌ" = "2", 
    "β" = "B",
    "ʤ" = "}\\\\textdyoghlig\\\\textipa{",
    "d͡ʒ" = "}\\\\textdyoghlig\\\\textipa{",
    "dʒ" = "}\\\\textdyoghlig\\\\textipa{", 
    "ð" = "D",
    "ə" = "@",
    "ɛ" = "E",
    "ɣ" = "G",
    "ɡ" = "g",
    "ɫ" = "\\|~l", 
    "ʎ" = "}\\\\textlambda\\\\textipa{",
    "ŋ" = "N",
    "ɲ" = "}\\\\textltailn\\\\textipa{", 
    "ɔ" = "O",
    "ɾ" = "R",
    "ʧ" = "}\\\\textteshlig\\\\textipa{", 
    "tʃ" = "}\\\\textteshlig\\\\textipa{", 
    "t͡ʃ" = "}\\\\textteshlig\\\\textipa{", 
    "θ" = "T", 
    "t͡s" = "}\\\\texttslig\\\\textipa{", 
    "ts" = "}\\\\texttslig\\\\textipa{", 
    "ʃ" = "S", 
    "ʒ" = "Z", 
    "ɪ" = "I", 
    "ʊ" = "U", 
    "ɜ" = "3", 
    "ɹ" = "\\*r", 
    "ː" = ":", 
    "ˈ" = '"'
  )
  tipa <- stringr::str_replace_all(ipa, dict)
  tipa <- paste0("\\textipa{", tipa, "}")
  return(tipa)
}

@@

tt(tbl_data) |> 
  style_tt(i = 0, bold = TRUE, line = "tb") |> 
  group_tt(i = list(
    "Experiment 1 (cat-ENG)" = 1,
    "Experiment 2 (spa-ENG)" = 6,
    "Experiment 3 (cat-SPA)" = 22)) |> 
  style_tt(i = 1, bold = TRUE, line = "b") |> 
  style_tt(i = 7, bold = TRUE, line = "tb") |> 
  style_tt(i = 24, bold = TRUE, line = "tb")
```
:::

::: {.content-visible when-format="docx"}

```{r tbl-surprises-docx}
#| label: tbl-surprises-docx
#| tbl-cap: "List of items with unexpectedly high accuracy: the Levenshtein similarityscore betwen the presented word (in Catalan or Spanish) and their correct English translation is zero, but participants, who are reportedly unfamiliar with the presented language, were on average >10% likely to guess the correct translation."
tbl_data <- list(
  "Experiment 1" = dataset_1,
  "Experiment 2" = dataset_2
) |>
  bind_rows(.id = "exp") |>
  mutate(
    translation = paste0(word_1, " - ",  word_2),
    ipa = paste0(ipa_1, " - ",  ipa_2),
  ) |> 
  summarise(
    n = n(),
    correct = sum(correct),
    .by = c(exp, group, lv, translation, ipa)
  ) |>
  mutate(
    accuracy = prop_adj(correct, n),
    accuracy_se = prop_adj_se(correct, n)
  ) |>
  dplyr::filter(lv == 0, accuracy >= 0.1) |>
  select(exp, translation, ipa, group, accuracy, accuracy_se) |>
  arrange(exp, group, desc(accuracy)) |> 
  select(translation, ipa, accuracy, accuracy_se) |> 
  mutate(accuracy = round(accuracy * 100, 2),
         accuracy_se = round(accuracy_se * 100, 2))

@@

:::

@@

Overall, participants reported prior knowledge more often for the Spanish words that showed unexpectedly high accuracy in Experiment 1 (see Discussion in Experiment 2) than for words with expected accuracy (see @fig-knowledge). Participants reported prior knowledge of Catalan words with unexpected accuracy in Experiment 1 as often as those with expected accuracy. This suggests that participants in Experiment 1 may have relied, to some extent, on their prior knowledge about form-meaning mappings to correctly translate some Spanish words. To isolate such an effect of prior Spanish knowledge, we run the same analysis as in Experiment 1 on the newly collected translations from Experiment 3, now excluding responses to words for which participants reported prior knowledge.

```{r fig-knowledge}
#| label: fig-knowledge
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Catalan and Spanish prior word knowledge, as reported by English native participants in Experiment 3. The X-axis indicates the proportion of participants that reported prior knowledge with the word. The Y-axis indicates participants' accuracy (the proportion of correct responses by participants). Each colour indicates whether the data point corresponds to a word that participants answered to with expected accuracy (red) or with unexpectedly high accuracy (blue), as defined by (no phonological similarity with the correct translation, by accuracy higher than 10%)."
dataset_3 |>
  mutate(surprise = translation %in% tbl_data$translation) |>
  summarise(
    knowledge = mean(knowledge),
    correct = sum(correct),
    n = n(),
    .by = c(translation, group, surprise)
  ) |>
  mutate(accuracy = prop_adj(correct, n),
  surprise = if_else(
    surprise,
    "Unexpected high accuracy\n(no phonological similarity)",
    "Expected accuracy"
  )) |> 
  ggplot(aes(knowledge, accuracy)) +
  facet_wrap(~group) +
  geom_point(
    stroke = 1, 
    size = 2, 
    alpha = 0.25, 
    color = RColorBrewer::brewer.pal(3, "Reds")[3], 
    shape = 1
  ) +
  labs(
    x = "Reported knowledge",
    y = "Accuracy"
  ) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
  )
```

@@

epreds_3 |> 
  ggplot(aes(lv_std, .epred)) +
  facet_grid(group ~ neigh_n_h_std) +
  stat_lineribbon(aes(fill_ramp = after_stat(level)),
    linewidth = 1 / 2,
    .width = c(0.95, 0.89, 0.78, 0.67, 0.50)
  ) +
  labs(
    x = "Similarity\n(Levenshtein similarity with correct translation)",
    y = "p(Correct)",
    colour = "Cross-language neighbourhood density",
    fill = "Cross-language neighbourhood density",
    linetype = "Cross-language neighbourhood density",
    fill_ramp = "CrI (%)"
  ) +
  scale_x_continuous(
    labels = \(x) scales::percent((x * sd(dataset_3$lv, na.rm = TRUE)) + mean(dataset_3$lv, na.rm = TRUE)), 
    breaks = (seq(0, 1, 0.2) - mean(dataset_3$lv, na.rm = TRUE)) / sd(dataset_3$lv, na.rm = TRUE)) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Reds") +
  theme(
    legend.position = "none",
    legend.box = "vertical"
  )

@@

Experiment 3 was a conceptual replication of Experiment 1 in which we gathered additional information about English native participants' prior familiarity with the meaning of the Catalan and Spanish words presented. After each trial, participants reported whether they had prior knowledge about the word meaning, and how confident they were about such information in a numeric scale from 0 to 7. To control for the effect of participants' prior knowledge abbout form-meaning mappings, we excluded from the analyses those words in which participants reported such prior knowledge. We found equivalent results to those in Experiment 1, indicating that participants' surprisingly good performance in the translation elicitation task was not due to being familiar with the meaning of words they were presented with.

>>>>

@@

@@

For each word pair in the Catalan and Spanish lists, we defined three predictors of interest: the lexical frequency of the correct English translation (*Frequency*), the phonological similarity between the presented word and its correct English translation (*Similarity*), and the presented word's number of cross-language phonological neighbours (*CLPN*) in English. *Frequency* was included as a nuisance predictor, under the hypothesis that---keeping other predictors constant---participants would provide correct responses more often when the correct translations are high frequency words in their native language. Lexical frequencies of correct translations were extracted from SUBTLEX-UK [@van2014subtlex], and transformed to Zipf scores. Translations without a lexical frequency value were excluded from data analysis (`r stim_exc_freq$cat_eng$n` in the Catalan list, `r stim_exc_freq$spa_eng$n` in the Spanish list).  *Similarity*, our main predictor of interest, was calculated as the Levenshtein similarity between the X-SAMPA transcriptions of each pair of translations using the `stringdist` R package [@van2014stringdist]. The Levenshtein distance computes the edit distance between two character strings---in this case, two phonological transcriptions---by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. We divided this edit distance by the number of characters in the longest X-SAMPA transcription of the translation pair. This transformation accounts for the fact that Levenshtein distance tends to increase with the length of the transcriptions. For interpretability, we subtracted this proportion from one, so that values closer to one correspond to higher similarity between phonological transcriptions, and values closer to zero correspond to lower similarity between phonological transcriptions. For example, the *table* (`teIb@l`)-*mesa* (`mesa`) translation pair had `r scales::percent(stringdist::stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (`trEIn`)-*tren* (`tREn`) translation pair had `r scales::percent(stringdist::stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity. We calculated the number of *CLPN* for each Catalan and Spanish word by counting the number of English words with same or higher lexical frequency, and whose phonological transcription (in X-SAMPA format) differed by up to one phoneme from that of the presented Catalan or Spanish word. Lexical frequencies and phonological transcriptions were extracted from the multilingual database CLEARPOND [@marian2012clearpond]^[Phonological transcriptions in CLEARPOND were generated from eSPEAK, [http://espeak.sourceforge.net/](http://espeak.sourceforge.net/)]]. @tbl-stimuli summarises the lexical frequency, *CLPN* and phonological overlap of the words included in the Catalan and the Spanish lists.

@@

$$
\begin{aligned}
&\textbf{Likelihood}  \\
y_{i} \sim & \text{Bernoulli}(p_{i}) \\ \\
&\textbf{Parameters}  \\
\text{Logit}(p_{i}) = &  \beta_{0[p,w]} + \beta_{1[p]} \text{Frequency}_{i} + \beta_{2[p]} \text{Similarity}_i +  \beta_{3[p]} \text{CLPN}_i + \beta_{4[p]} \text{Group}_i + \\
&\beta_{5[p]} (\text{Similarity}_i \times \text{CLPN}_i) + \beta_{6[p]} (\text{Similarity}_i \times \text{Group}_i) + \\
& \beta_{7[p]} (\text{CLPN}_i \times \text{Group}_i) + \beta_{8[p]} (\text{Similarity}_i \times \text{CLPN}_i \times \text{Group}_i) \\ \\
\beta_{0-8[p,w]} \sim & \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \text{ and  word } w \text{ in 1, ..., } W \\
\beta_{1-8[p]} \sim &  \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \\ \\
&\textbf{Prior}  \\
\mu_{\beta_{p,w}}  \sim &  \mathcal{N}(0, 0.1) \\
\sigma_{\beta_{p}},  \sigma_{\beta_{w}} \sim & \text{HalfCauchy}(0, 0.1) \\
\rho_{p}, \rho_{w} \sim & \text{LKJCorr}(8) \\
\end{aligned}
$$ {#eq-1}

@@

All analyses were performed in the R environment [@rcoreteam2013language]. We used the tidyverse family of R packages [@wickham2019welcome] to process data and to generate figures. We used the `brms` R package [@burkner2017brms] using the `cmdstanr` back-end to the Stan probabilistic language [@carpenter2017stan] to estimate and compare the models (see Appendix 1 for model diagnostics). Numeric predictors were standardised before entering the model by subtracting the mean and dividing by the standard deviation.

@@

tbl_data |>
  gt(
    rowname_col = "group",
    groupname_col = "exp"
  ) |>
  fmt_number(is.numeric, decimals = 2) |>
  fmt_integer(c(matches("sum|min|max"), n_participants)) |>
  fmt_number(matches("correct"), scale_by = 100) |>
  tab_spanner("Accuracy (%)", matches("correct")) |>
  tab_spanner("Valid trials", matches("n_trials")) |>
  cols_merge_range(n_trials_min, n_trials_max) |>
  cols_merge_range(correct_min, correct_max) |>
  cols_label(
    n_trials_sum = "N trials",
    n_participants = "N",
    n_trials_mean = "Mean",
    n_trials_sd = "SD",
    n_trials_min = "Range",
    correct_mean = "Mean",
    correct_sd = "SD",
    correct_sem = "SE",
    correct_min = "Range"
  ) |>
  tab_style(
    cell_text(align = "center"),
    cells_column_labels()
  ) |>
  tab_style(
    cell_text(style = "italic"),
    cells_column_labels()
  ) |>
  tab_style(
    cell_text(weight = "bold"),
    cells_column_spanners()
  ) |>
  tab_style(
    cell_text(align = "left"),
    list(cells_body(), cells_column_labels())
  )
```

@@

epreds_1 |> 
  ggplot(aes(lv_std, .epred)) +
  facet_grid(group ~ neigh_n_h_std) +
  stat_lineribbon(aes(fill_ramp = after_stat(level)),
    linewidth = 1 / 2,
    .width = c(0.95, 0.89, 0.78, 0.67, 0.50)
  ) +
  labs(
    x = "Similarity\n(Levenshtein similarity with correct translation)",
    y = "p(Correct)",
    colour = "Cross-language Phonological Neighbours",
    fill = "Cross-language Phonological Neighbours",
    linetype = "Cross-language Phonological Neighbours",
    fill_ramp = "CrI (%)"
  ) +
  scale_x_continuous(
    labels = \(x) scales::percent((x * sd(dataset_1$lv, na.rm = TRUE)) + mean(dataset_1$lv, na.rm = TRUE)), 
    breaks = (seq(0, 1, 0.2) - mean(dataset_1$lv, na.rm = TRUE)) / sd(dataset_1$lv, na.rm = TRUE)) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Reds") +
  theme(
    legend.position = "none",
    legend.box = "vertical"
  )
```

@@

epreds_2 |> 
  ggplot(aes(lv_std, .epred)) +
  facet_grid(~ neigh_n_h_std) +
  stat_lineribbon(aes(fill_ramp = after_stat(level)),
    linewidth = 1 / 2,
    .width = c(0.95, 0.89, 0.78, 0.67, 0.50)
  ) +
  labs(
    x = "Similarity\n(Levenshtein similarity with correct translation)",
    y = "p(Correct)",
    colour = "Cross-language Phonological Neighbours",
    fill = "Cross-language Phonological Neighbours",
    linetype = "Cross-language Phonological Neighbours",
    fill_ramp = "CrI (%)"
  ) +
  scale_x_continuous(
    labels = \(x) scales::percent((x * sd(dataset_2$lv, na.rm = TRUE)) + mean(dataset_2$lv, na.rm = TRUE)), 
    breaks = (seq(0, 1, 0.2) - mean(dataset_2$lv, na.rm = TRUE)) / sd(dataset_2$lv, na.rm = TRUE)) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Reds") +
  theme(
    legend.position = "none",
    legend.box = "vertical"
  )
```

@@

Experiment 2 was an extension of Experiment 1 to a population of monolinguals whose native language is typologically close to the presented language. We presented Catalan words to Spanish native adults who were reportedly unfamiliar with Catalan. Our results indicate a similar pattern of results as those in Experiment 1: participants were able to provide correct translations of presented Catalan words, provided that the Catalan words shared some degree of phonological similarity with their Spanish translation, and that the number of phonological neighbours with higher lexical frequency was reduced. In contrast with the results in Experiment 1, the positive impact of phonological similarity on participants' performance in Experiment 2 larger. Spanish natives in Experiment 2 exploited phonological similarity more strongly than English natives in Experiment 1. Overall, this suggests that participants in Experiment 2, who were natives of a typologically similar language (Spanish) to the presented language (Catalan), benefited more strongly from phonological similarity than participants in Experiment 1, who were natives of a typologically less similar language (English) to the presented language (Catalan, Spanish). Participants from both Experiment 1 and 2 benefited strongly from phonological similarity to correctly translate words from a non-native, reportedly unfamiliar language. This pattern of results holds for most of the presented stimuli, but some low-similarity Catalan and Spanish words were responded to surprisingly accurately by English listeners. Given that participants were reportedly unfamiliar with both languages, it was expected that participants would be very unlikely to provide correct translations for words sharing little to no phonological similarity to their correct translation. @tbl-surprises shows a list of Catalan and Spanish words to which participants provided responses with $\geq$ 10 average accuracy.

```{r}
#| label: tbl-surprises
#| tbl-cap: "List of items with unexpectedly high accuracy: the Levenshtein similarity score between the presented word (in Catalan or Spanish) and their correct English translation is zero, but participants, who are reportedly unfamiliar with the presented language, were on average >10% likely to guess the correct translation."
ipa_to_tipa <- function(ipa) {
  dict <- c(
    "ɑ" = "A", 
    "ɒ" = "6",
    "æ" = "\\ae", 
    "ʌ" = "2", 
    "β" = "B",
    "ʤ" = "}\\\\textdyoghlig\\\\textipa{",
    "d͡ʒ" = "}\\\\textdyoghlig\\\\textipa{",
    "dʒ" = "}\\\\textdyoghlig\\\\textipa{", 
    "ð" = "D",
    "ə" = "@",
    "ɛ" = "E",
    "ɣ" = "G",
    "ɡ" = "g",
    "ɫ" = "\\|~l", 
    "ʎ" = "}\\\\textlambda\\\\textipa{",
    "ŋ" = "N",
    "ɲ" = "}\\\\textltailn\\\\textipa{", 
    "ɔ" = "O",
    "ɾ" = "R",
    "ʧ" = "}\\\\textteshlig\\\\textipa{", 
    "tʃ" = "}\\\\textteshlig\\\\textipa{", 
    "t͡ʃ" = "}\\\\textteshlig\\\\textipa{", 
    "θ" = "T", 
    "t͡s" = "}\\\\texttslig\\\\textipa{", 
    "ts" = "}\\\\texttslig\\\\textipa{", 
    "ʃ" = "S", 
    "ʒ" = "Z", 
    "ɪ" = "I", 
    "ʊ" = "U", 
    "ɜ" = "3", 
    "ɹ" = "\\*r", 
    "ː" = ":", 
    "ˈ" = '"'
  )
  tipa <- stringr::str_replace_all(ipa, dict)
  tipa <- paste0("\\textipa{", tipa, "}")
  return(tipa)
}

@@

@@

@@

Overall, participants reported prior knowledge more often for the Spanish words that showed unexpectedly high accuracy in Experiment 1 (see Discussion in Experiment 2) than for words with expected accuracy (see @fig-knowledge). Participants reported prior knowledge of Catalan words with unexpected accuracy in Experiment 1 as often as those with expected accuracy. This suggests that participants in Experiment 1 may have relied, to some extent, on their prior knowledge about form-meaning mappings to correctly translate some Spanish words. To isolate such an effect of prior Spanish knowledge, we ran the same analysis as in Experiment 1 on the newly collected translations from Experiment 3, now excluding responses to words for which participants reported prior knowledge.

```{r fig-knowledge}
#| label: fig-knowledge
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Catalan and Spanish prior word knowledge reported by English native participants in Experiment 3. The X-axis indicates the proportion of participants that reported prior knowledge with the word. The Y-axis indicates participants' accuracy (the proportion of participants that translated the word correctly). For visualisation purposes, data points have been aggregated so that numbers and color coding show the number of data points with identical accuracy and proportion of reported knowledge."

bin_interval <- seq(0, 1, 0.05)
dataset_3 |>
  mutate(surprise = translation %in% tbl_data$Translation) |>
  summarise(
    knowledge = mean(knowledge),
    correct = sum(correct),
    n = n(),
    .by = c(translation, group, surprise)
  ) |>
  mutate(
    accuracy = prop_adj(correct, n),
  accuracy_binned = cut(accuracy, bin_interval, labels = bin_interval[-length(bin_interval)], include.lowest = TRUE, right = TRUE),
    knowledge_binned = cut(knowledge, bin_interval, labels = as.numeric(bin_interval[-length(bin_interval)]), include.lowest = TRUE, right = TRUE)
  ) |> 
    summarise(
      n = n(),
      .by = c(accuracy_binned, knowledge_binned, group)
    ) |> 
      mutate(is_gt_20 = n > 20,
      knowledge_binned = as.double(as.character(knowledge_binned)),
      accuracy_binned = as.double(as.character(accuracy_binned))) |> 
  ggplot(aes(knowledge_binned, accuracy_binned, fill = n, label = n, color = is_gt_20)) +
  facet_wrap(~group) +
  geom_tile(color = NA) +
  geom_text(size = 3) +
  labs(
    x = "Reported knowledge",
    y = "Accuracy"
  ) +
  scale_fill_distiller(palette = "Reds", direction=1, limits = c(0, 37)) +
  scale_color_manual(values = c("black", "white")) +
  scale_x_continuous(labels = scales::percent, limits = c(-0.05, 1)) +
  scale_y_continuous(labels = scales::percent, limits = c(-0.05, 1), breaks = seq(0, 1, 0.1)) +
  theme(
    legend.position = "none",
  )
```

@@

epreds_3 |> 
  ggplot(aes(lv_std, .epred)) +
  facet_grid(group ~ neigh_n_h_std) +
  stat_lineribbon(aes(fill_ramp = after_stat(level)),
    linewidth = 1 / 2,
    .width = c(0.95, 0.89, 0.78, 0.67, 0.50)
  ) +
  labs(
    x = "Similarity\n(Levenshtein similarity with correct translation)",
    y = "p(Correct)",
    colour = "Cross-language Phonological Neighbours",
    fill = "Cross-language Phonological Neighbours",
    linetype = "Cross-language Phonological Neighbours",
    fill_ramp = "CrI (%)"
  ) +
  scale_x_continuous(
    labels = \(x) scales::percent((x * sd(dataset_3$lv, na.rm = TRUE)) + mean(dataset_3$lv, na.rm = TRUE)), 
    breaks = (seq(0, 1, 0.2) - mean(dataset_3$lv, na.rm = TRUE)) / sd(dataset_3$lv, na.rm = TRUE)) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Reds") +
  theme(
    legend.position = "none",
    legend.box = "vertical"
  )

@@

Experiment 3 was a conceptual replication of Experiment 1 in which we gathered additional information about English native participants' prior familiarity with the meaning of the Catalan and Spanish words presented. After each trial, participants reported whether they had prior knowledge about the word meaning, and how confident they were about such information in a numeric scale from 0 to 7. To control for the effect of participants' prior knowledge about form-meaning mappings, we excluded from the analyses those words in which participants reported such prior knowledge. We found equivalent results to those in Experiment 1, indicating that participants' surprisingly good performance in the translation elicitation task was not due to being familiar with the meaning of words they were presented with.
